Compositional Quantitative Resource Analysis of C Programs

### 2014-06-20 paper/main.tex
minor changes

that works automatically, efficiently and precisely for many examples,
can be used interactively if the automation fails, and produces

@@

analysis enable semi-automatic interactive resource bound development
in which an automatically derived bound can be used in the
quantitative logic and a hand derived bound can be used by the
automatic analysis.  Both the automatic amortized analysis and the
quantitative logic are naturally compositional and describe the
resource behavior of code fragments without referring to the source
code.  As a result, we naturally derive global bounds that are
functions of the input parameters of the program.  To the best of our
knowledge, our system is the only one that can deal with resource like
memory that may become available during the execution.

@@

We have evaluated the quantitative logic and the automatic analysis
with system code and examples from the literature. \pref{app:cat}
contains a $50$ (XXX) challenging loop and recursion patterns that we
collected from open source software and the literature.  Our analysis
can find asymptotically tight bounds for $48$ of these patterns and in
most cases the derived constant factors are tight.  We compared our
automatic analyzer to other resource-analysis systems.  However, our
analyzer and Rank are currently the only publically available tools
that can directly operate on C programs.  We selected $5$ example for
which we derived linear bounds and translated them to term-rewriting
systems by hand in order to run them with the tools KoAT and Pubs.
Both tools and Rank were not able to find linear bounds for the
examples.  We contacted the authors of LOOPUS and SPEED but were not
able to obtain a tool to perform a comparison.

@@

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{L:Assert}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q \implies \code{istrue}~\sem{e}_\state }
  { \code{assert}~e }
  { Q }
}
%
\and
\Rule{L:Skip}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_s + Q }{ \code{skip} }{ Q }
}
%
\and
\Rule{L:Break}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_b + B }{ \code{break} }{ Q }
}
%
\and
\Rule{L:Return}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_r + R\,x }{ \code{return}~x }{ Q }
}
%
\and
\Rule{L:Update}
{ }
{ \Delta; B; R \vdash
\htriple
  { \lambda \state.\,M_u + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
}
%
\and
\Rule{L:Call}
{
  \Delta f = \forall z\,\vec v.(P_f\,z\,\vec v, Q_f\,z) \\
  P \models P_f \, y \, (\state(\vec x)) \land A \\
  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
}
{ \Delta; B; R \vdash
\htriple
  { M_f + P }
  { r \gets f(\vec x) }
  { Q }
}
%
\and
\Rule{L:Loop}
{ \Delta; Q; R \vdash
\htriple
  { I }{ S }{ I + M_l }
}
{ \Delta; B; R \vdash
\htriple
  { I }{ \code{loop}~S }{ Q }
}
%
\and
\Rule{L:Seq}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1 }{ Q' + M_s }
\\
  \Delta; B; R \vdash
\htriple{ Q' }{ S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1;S_2 }{ Q }
}
%
\and
\Rule{L:If}
{ \Delta; B; R \vdash
\htriple
  { P \land \code{istrue}~\sem{e}_\state - M_c^1}
  { S_1 }{ Q }
\\
  \Delta; B; R \vdash
\htriple
  { P \land \code{isfalse}~\sem{e}_\state - M_c^2 }
  { S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple
  { P }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
}
%
\and
\Rule{L:Weaken}
{
  P \models P' \\
  \Delta; B'; R' \vdash
  \htriple{ P' }{ S }{ Q' } \\
  Q' \models Q \\
  B' \models B \\
  \forall v.\, (R'\,v \models R\,v)
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\and
\Rule{L:Frame}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q } \\
  x \in \mathbb Q_0^+
}
{ \Delta; B + x; R + x \vdash
\htriple{ P + x }{ S }{ Q + x }
}
%
\and
\Rule{L:Extend}
{ \forall f\,P_f\,Q_f.\,
  \Delta'f = \forall z\,\vec v.(P_f\,z\,\vec v, Q_f\,z)
  \rightarrow
  \forall y\,\vec v.\,
  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
\\
  \Delta \cup \Delta'; B; R \vdash
  \htriple{ P }{ S }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\end{mathpar}

\caption{Rules of the Quantitative Hoare Logic}
\end{figure*}

>>>>

that is based on a solid semantic foundation. It works automatically,
efficiently and precisely for many examples, can be used interactively
for manual bound derivation if the automation fails, and produces

@@

analysis enable semi-automatic interactive resource-bound development.
An automatically derived bound can be used in the quantitative logic
and a hand derived bound can be used by the automatic analysis.  Both
the automatic amortized analysis and the quantitative logic are
naturally compositional and describe the resource behavior of code
fragments without referring to the source code.  As a result, we
naturally derive global bounds that are functions of the input
parameters of the program.  To the best of our knowledge, our system
is the only one that can deal with resource like memory that may
become available during the execution.

@@

We have evaluated the quantitative logic and the automatic analysis
with system code and examples from the literature. \pref{app:cat}
contains more than $30$ (XXX) challenging loop and recursion patterns
that we collected from open source software and the literature.  Our
analysis can find asymptotically tight bounds for $30$ all but 3 of
these patterns and in most cases the derived constant factors are
tight.  We compared our automatic analyzer to other resource-analysis
systems.  We selected $5$ examples (see Section XXX) for which we
derived linear bounds and translated them to term-rewriting systems by
hand in order to run them with the tools KoAT and Pubs.  Both tools
and Rank were not able to find linear bounds for the examples.  LOOPUS
is able to derive a bound on XXX of the examples.  But in contrast to
our analysis, the constant factors are not tight.  We also contacted
the authors of SPEED but were not able to obtain a tool to perform a
comparison.

@@

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{L:Assert}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q \implies \code{istrue}~\sem{e}_\state + M_a }
  { \code{assert}~e }
  { Q }
}
%
\and
\Rule{L:Skip}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_s + Q }{ \code{skip} }{ Q }
}
%
\and
\Rule{L:Break}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_b + B }{ \code{break} }{ Q }
}
%
\and
\Rule{L:Return}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_r + R\,x }{ \code{return}~x }{ Q }
}
%
\and
\Rule{L:Update}
{ }
{ \Delta; B; R \vdash
\htriple
  { \lambda \state.\,M_u + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
}
%
\and
\Rule{L:Call}
{
  \Delta f = \forall z\,\vec v.(P_f\,z\,\vec v, Q_f\,z) \\
  P \models P_f \, y \, (\state(\vec x)) \land A \\
  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
}
{ \Delta; B; R \vdash
\htriple
  { M_f + P }
  { r \gets f(\vec x) }
  { Q }
}
%
\and
\Rule{L:Loop}
{ \Delta; Q; R \vdash
\htriple
  { I }{ S }{ I + M_l }
}
{ \Delta; B; R \vdash
\htriple
  { I }{ \code{loop}~S }{ Q }
}
%
\and
\Rule{L:Seq}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1 }{ Q' + M_s }
\\
  \Delta; B; R \vdash
\htriple{ Q' }{ S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1;S_2 }{ Q }
}
%
\and
\Rule{L:If}
{ \Delta; B; R \vdash
\htriple
  { P \land \code{istrue}~\sem{e}_\state - M_c^1}
  { S_1 }{ Q }
\\
  \Delta; B; R \vdash
\htriple
  { P \land \code{isfalse}~\sem{e}_\state - M_c^2 }
  { S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple
  { P }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
}
%
\and
\Rule{L:Weaken}
{
  P \models P' \\
  \Delta; B'; R' \vdash
  \htriple{ P' }{ S }{ Q' } \\
  Q' \models Q \\
  B' \models B \\
  \forall v.\, (R'\,v \models R\,v)
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\and
\Rule{L:Frame}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q } \\
  x \in \mathbb Q_0^+
}
{ \Delta; B + x; R + x \vdash
\htriple{ P + x }{ S }{ Q + x }
}
%
\and
\Rule{L:Extend}
{ \forall f\,P_f\,Q_f.\,
  \Delta'f = \forall z\,\vec v.(P_f\,z\,\vec v, Q_f\,z)
  \rightarrow
  \forall y\,\vec v.\,
  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
\\
  \Delta \cup \Delta'; B; R \vdash
  \htriple{ P }{ S }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\end{mathpar}

\caption{Rules of the Quantitative Hoare Logic}
\label{fig:logic}
\end{figure*}

\begin{figure}
\begin{lstlisting}
${\color{blue} \{ A(0, N) + N \} }$
i=1; m=a[0];
${\color{blue}%
  \{ (i {=} 1 \land m=a[0]) + A(1, N) + (N-i) \} }$
while (i<N) {
  ${\color{blue}%
    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i, N) + (N-i) \} }$
  if (a[i] > m)
    m = a[i];
  ${\color{blue}%
    \{ (m = \max_{k \in [0, i]} a[k]) + A(i+1, N) + (N-i) \} }$
  i = i+1;
  ${\color{blue}%
    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i, N) + (N-i) \} }$
}
${\color{blue} \{ 0 \} }$
\end{lstlisting}
\caption{Example derivation where we noted $A(i, N)$
  for $\#\{ k \mid k \ge i \land \forall\, 0\le j<k.\, a[j] < a[k]\}$}
\label{fig:exemple}
\end{figure}
